MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-24 Manuscript/Manuscript-2.Rmd
vsn 2

First, it presumes that values of $\lambda$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. Therefore, understanding the precision in estimating $\lambda$ is paramount. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise with greater sample sizes [@Cohen1988]. However, this conclusion also assumes that the precision in estimating $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while estimates of $\lambda$ are within a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets [e.g., meta-analysis: @HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared across datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all model parameters and test statistics are effect sizes, and thus many summary measures must first be converted to standardized units (i.e., an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is a statistical effect size can comparisons of estimates across datasets be interpretable. For the case of $\lambda$, this has not yet been explored.  \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that the precision in estimating $\lambda$ is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Additionally, the same $\lambda_{est}$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Thus, $\lambda$ is not a reliable effect size for measuring the strength of phylogenetic signal. We subsequently derive a standardized effect size for measuring the strength of phylogenetic signal in phenotypic datasets, and apply the concept to two common measures of phylogenetic signal: $\lambda$ and $Kappa$. Through simulations across a wide range of conditions, we find that the precision of effect sizes based on $\lambda$ ($Z_{\lambda}$) are less reliable than that those based on $Kappa$ ($Z_K$), implying that $Z_K$ is a more robust effect size measure. Additionally, we propose a two-sample test statistic that may be used to compare the strength of phylogenetic signal among datasets. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

@@

## The Precision of $\lambda$ is Variable

We conducted a series of computer simulations to evaluate the precision of Pagel's $\lambda$. Our primary simulations were based on pure-birth phylogenies; however, we also evaluated patterns on both balanced and pectinate trees to determine whether tree shape affected our findings (see Supporting Information). First we generated 50 pure-birth phylogenies at each of six different tree sizes, ranging from 32 to 1024 taxa ($n=2^5 - 2^{10}$). Next, we rescaled the simulated phylogenies by multiplying the internal branches by $\lambda_{in} = 0.0 \to 1.0$ using 21 intervals of 0.05 units, resulting in 1050 scaled phylogenies at each level of species richness ($n$). Continuous traits were then simulated on each phylogeny under a Brownian motion model of evolution to obtain datasets with differing levels of phylogenetic signal, that ranged from no phylogenetic signal (when $\lambda_{in} =0$), to phylogenetic signal corresponding reflecting Brownian motion (when $\lambda_{in} =1$). We then calculated the precision (variance: $\sigma^2_\lambda$) of $\lambda$ at each input level of phylogenetic signal and at each level of species richness ($n$). \hfill\break

We also evaluated the precision of $\lambda$ when estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$). Here, an independent variable $X$ was simulated on each phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$. Here, the association between $Y$ and $X$  was modeled using a range of values: $\beta=(0.0,0.25, 0.5, 0.75,1.0)$, and the residual error was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution: $\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})$: [see @Revell2010 for a similar simulation design]. The fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained. Precision estimates ($\sigma^2_\lambda$) at each input level of phylogenetic signal and at each level of species richness ($n$) were then observed. All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supporting Information. \hfill\break

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 1). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal, with the worst precision at intermediate levels of signal ($\lambda_{in}\approx0.5$), and improved precision as input levels approaches 0 and 1 (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). Results were similar when $\lambda$ was co-estimated with regression parameters in PGLS regression (Figure 2). Here, regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell 2010 [-@Revell2010] (see Supporting Information). However, estimates of phylogenetic signal were not, and the spread of $\lambda_{est}$ was even broader tham that observed when $\lambda$ was estimated for only the dependent variable. Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize observed levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

The above findings have several important implications for empirical studies.

-Empirical examples: how interpretation can get folks ito trouble.

## A Standardized Effect Size for Phylogenetic Signal

  -Derivation of effect size (expand on Adams and Collyer 2019 suggestion that an effect size for phylogenetic signal is needed. The ‘gold standard’ of these is a standaredized effect size: (following the meta-analytic literature beginning with Glass 1976).  In other words, a Z-score. So, how does Z work when based on lambda? On kappa?
  - Z-score. could be Lambda or Kappa.  Show it is Kappa
-Comparing the strength of physig
  - two sample Z-score

@@

Finally, for comparison we characterized the strength of phylogenetic signal in each dataset using a standardized effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] based on $Kappa$. As suggested by Adams and Collyer [-@AdamsCollyer2019], an effect size for phylogenetic signal may be estimated as: $Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}$, where $K_{obs}$ was the observed $Kappa$, and $\mu_K$ and $\sigma_K$ were the mean and standard deviation of the empirical sampling distribution of values obtained from the permutation distribution. $Z_K$ describes the strength of phylogenetic signal as a standard deviate from its sampling distribution, and thus directly measures the strength of signal in a manner that is comparable across datasets. Variation in the set of $Z_K$ at each input level of phylogenetic signal was then calculated as an estimate of precision in $Z_K$. However, because $Z_K$ differs in scale from $\lambda$, we used a linear normalization to standardize $Z_K$ to a uniform distribution ($0\rightarrow1$), and estimated the precision of $Z_K$ from the normalized values. \hfill\break

>>>>

First, it presumes that values of $\lambda$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. Therefore, understanding the precision in estimating $\lambda$ is paramount. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise at greater sample sizes [@Cohen1988]. However, this conclusion also assumes that the precision of $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while estimates of $\lambda$ are within a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets as in meta-analysis [@HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared among datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all model parameters and test statistics are effect sizes, and thus many summary measures must first be converted to standardized units (i.e., an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is a statistical effect size can comparisons of estimates across datasets be interpretable. For the case of $\lambda$, this has not yet been explored.  \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that the precision in estimating $\lambda$ is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Additionally, the same $\lambda_{est}$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Thus, $\lambda$ is not a reliable estimate of the strength of phylogenetic signal in phenotypic data. We subsequently derive a standardized effect size for measuring the strength of phylogenetic signal in phenotypic datasets, and apply the concept to two common measures of phylogenetic signal: $\lambda$ and $Kappa$. Through simulations across a wide range of conditions, we find that the precision of effect sizes based on $\lambda$ ($Z_{\lambda}$) are less reliable than that those based on $Kappa$ ($Z_K$), implying that $Z_K$ is a more robust effect size measure. Additionally, we propose a two-sample test statistic that may be used to compare the strength of phylogenetic signal among datasets, and provide an empirical example to demonstrate its use. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

@@

## *The Precision of $\lambda$ is Variable*

We conducted a series of computer simulations to evaluate the precision of Pagel's $\lambda$. Our primary simulations were based on pure-birth phylogenies; however, we also evaluated patterns on both balanced and pectinate trees to determine whether tree shape affected our findings (see Supporting Information). First we generated 50 pure-birth phylogenies at each of six different tree sizes, ranging from 32 to 1024 taxa ($n=2^5 - 2^{10}$). Next, we rescaled the simulated phylogenies by multiplying the internal branches by $\lambda_{in}$, using 21 intervals of 0.05 units across its range ($\lambda_{in} = 0.0 \to 1.0$), resulting in 1050 scaled phylogenies at each level of species richness ($n$). Continuous traits were then simulated on each phylogeny under a Brownian motion model of evolution to obtain datasets with differing levels of phylogenetic signal, that ranged from no phylogenetic signal (when $\lambda_{in} =0$), to phylogenetic signal corresponding reflecting Brownian motion (when $\lambda_{in} =1$). For each dataset we then estimated phylogenetic signal ($\lambda_{est}$), and calculated the precision of $\lambda$ using the variance ($\sigma^2_\lambda$) across datasets at each input level of phylogenetic signal and level of species richness. \hfill\break

We also evaluated the precision of $\lambda$ when estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$). Here, an independent variable $X$ was simulated on each phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$. Here, the association between $Y$ and $X$  was modeled using a range of values: $\beta=(0.0,0.25, 0.5, 0.75,1.0)$, and the residual error was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution: $\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})$: [see @Revell2010 for a similar simulation design]. The fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained. Precision estimates ($\sigma^2_\lambda$) at each input level of phylogenetic signal and level of species richness were then observed. \hfill\break

All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supporting Information. \hfill\break

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 1). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal, with the worst precision at intermediate levels of signal ($\lambda_{in}\approx0.5$), and improved precision as input levels approached the extremes of its range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). Results were similar when $\lambda$ was co-estimated with regression parameters in PGLS regression (Figure 2). Here, regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell 2010 [-@Revell2010] (see Supporting Information). However, estimates of phylogenetic signal were not, and the spread of $\lambda_{est}$ was even broader than that observed when $\lambda$ was estimated for only the dependent variable. Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize observed levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

## *A Standardized Effect Size for Phylogenetic Signal*

The results above demonstate that $\lambda$ is not a reliable estimate of the phylogenetic signal in phenotypic data. As such, biological interpretations of the strength of phylogenetic signal, and comparisons of the magnitude of such effects across datasets, are severely compromised when based on this parameter. As an alternative, we propose that summary estimates of phylogenetic signal be converted to effect sizes for these purposes. A standardized effect size is found as:

\begin{align}
    Z_{\theta}=\frac{\theta_{obs}-E(\theta)}{\sigma_\theta}
\end{align}

where $\theta_{obs}$ is the observed test statistic, $E(\theta)$ is its expected value under the null hypothesis, and $\sigma_\theta$ is its standard error [@Glass1976; @Cohen1988;@Rosenthal1994]. $Z_{\theta}$ expresses the magnitude of the effect in $\theta_{obs}$ by transforming the original test statistic to a standard normal deviate [@Glass1976; @Kelley2012]. Here, $\theta_{obs}$ and $\sigma_\theta$ are estimated from the data, while $E(\theta)$ is obtained from the distribution of $\theta$ derived from parametric theory. However, recent advances in resampling theory [@Collyer_et_al2015a;@AdamsCollyer2016; @AdamsCollyer2019b] have shown that $E(\theta)$ and $\sigma_\theta$ may also be obtained from an empirical sampling distribution obtained from permutation procedures. \hfill\break

Adams and Collyer [-@AdamsCollyer2019] recently suggested that the strength of phylogenetic signal could be represented an effect size, obtained from $Kappa$ and its empirical sampling distribution from permutation. Here we formalize that suggestion, and find an effect size as:

\begin{align}
    Z_{K}=\frac{K_{obs}-\hat\mu_K}{\hat\sigma_K}
\end{align}

Similarly, an effect size based on $\lambda$ could be envisioned as:

\begin{align}
    Z_{\lambda}=\frac{\lambda_{obs}-0}{\hat\sigma_\lambda}
\end{align}

Note that under the null hypothesis, $E(\lambda)=0$, a no phylogenetic signal is expected under this condition [@Freckleton_et_al2002].

  - Z-score. could be Lambda or Kappa.  Show it is Kappa
-Comparing the strength of physig
  - two sample Z-score

@@

Finally, for comparison we characterized the strength of phylogenetic signal in each dataset using a standardized effect size [$Z_K$: sensu @AdamsCollyer2016; @AdamsCollyer2019b] based on $Kappa$.

Variation in the set of $Z_K$ at each input level of phylogenetic signal was then calculated as an estimate of precision in $Z_K$. However, because $Z_K$ differs in scale from $\lambda$, we used a linear normalization to standardize $Z_K$ to a uniform distribution ($0\rightarrow1$), and estimated the precision of $Z_K$ from the normalized values. \hfill\break
